Galaxy image statistical generator: how to be confident?

### 2024-09-11 paper/main.tex
experiment

\begin{document}
\begin{nouppercase}
\maketitle
\end{nouppercase}
%\renewcommand{\baselinestretch}{0.75}\normalsize
%\tableofcontents
%\renewcommand{\baselinestretch}{1.0}\normalsize
%
\begin{abstract}
blabla
\\
\smallskip
\noindent \textbf{Keywords.} diffusion models, U-Net
\end{abstract}

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}. Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014}\footnote{nb. notice that \texttt{arxiv} paper have been updated in 2022.} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep[e.g.]{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.
%
\section{Generative Models}
%
{\color{red} Mention ergodic vs non ergodic, reference Allys et al. Lempereur et al. simulation of ergodic case.}\hfill\\

The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.  It is also assumed that the data are i.i.d samples of $p(\bm{x})$.  In the following, it is not the intention to describe in high details the different  strategies and their implementations but rather {\color{red} to give some guide lines (à modifier)}.
%
\subsection{Variational Auto-Encoder}
%
Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$. In general the latent space has a much lower dimensionality than the data space which imply a information compression. Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$. However, the likelihood  $p(\bm{x}|\bm{z})$ is also unknown which leads to the introduction of a parametrized version $p_{\bm{\theta_2}}(\bm{x}|\bm{z})$ with the differentiability properties as $\pi_{\bm{\theta}_1}$. Using the notation $\bm{\theta}=(\bm{\theta_1},\bm{\theta_2})$,  these parameters are adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 
$\int p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z}) dz$ \citep{Kingma2014}. To overcome such problem, one uses first the Bayes relation
\begin{equation}
p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})
\end{equation}
and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$. If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{\bm{x}\sim p}[\log(p(\bm{x})/q(\bm{x}))]$, then taking the expected value according\footnote{We represent the sampling of variates $\bm{x}$ from a
distribution $p(\bm{x})$ using the notation $\bm{x}\sim p(\bm{x})$.} to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
\begin{align}
\log p_{\bm{\theta}}(x) &= \log p_{\bm{\theta}_2}(\bm{x}|\bm{z}) + \log \pi_{\bm{\theta}_1}(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
\end{align}
yields 
\begin{align}
\log p_{\bm{\theta}}(x)  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
\label{eq-VAE-ELBO}
\end{align}
where we have emphasized the \textit{evidence lower bound} (aka ELBO)
\begin{equation}
\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z}))
\end{equation}
and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$. However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary  variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.

@@

An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. They used \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size in the range $10^4-10^5$.

So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}

>>>>

\begin{document}
\begin{nouppercase}
\maketitle
\end{nouppercase}
%\renewcommand{\baselinestretch}{0.75}\normalsize
%\tableofcontents
%\renewcommand{\baselinestretch}{1.0}\normalsize
%
\begin{abstract}
Todo...
\\
\smallskip
\noindent \textbf{Keywords.} diffusion score based model, normalizing flows, U-Net denoiser, Glow
\end{abstract}

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task especially in the context of non-ergodic processes that has made a breakthrough in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}. Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014}\footnote{nb. notice that \texttt{arxiv} paper have been updated in 2022.} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep[e.g.]{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.
%
\section{Generative Models}
%
The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.  It is also assumed that the data are i.i.d samples of $p(\bm{x})$.  In the following sections, the intention is more to give some background knowledge rather than describe in great details the different  architecture types and their implementations.
%
\subsection{Variational Auto-Encoder}
%
Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$. In general the latent space has a much lower dimensionality than the data space which imply a information compression. Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$. However, the likelihood  $p(\bm{x}|\bm{z})$ is also unknown which leads to the introduction of a parametrized version $p_{\bm{\theta_2}}(\bm{x}|\bm{z})$ with the differentiability properties as $\pi_{\bm{\theta}_1}$. Using the notation $\bm{\theta}=(\bm{\theta_1},\bm{\theta_2})$,  these parameters are adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 
$\int p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z}) dz$ \citep{Kingma2014}. To overcome such problem, one uses first the Bayes relation
\begin{equation}
p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})
\end{equation}
and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$. If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{\bm{x}\sim p}[\log(p(\bm{x})/q(\bm{x}))]$, then taking the expected value according\footnote{We represent the sampling of variates $\bm{x}$ from a
distribution $p(\bm{x})$ using the notation $\bm{x}\sim p(\bm{x})$.} to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
\begin{align}
\log p_{\bm{\theta}}(x) &= \log p_{\bm{\theta}_2}(\bm{x}|\bm{z}) + \log \pi_{\bm{\theta}_1}(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
\end{align}
yields 
\begin{align}
\log p_{\bm{\theta}}(x)  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
\label{eq-VAE-ELBO}
\end{align}
where we have emphasized the \textit{evidence lower bound} (aka ELBO)
\begin{equation}
\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z}))
\end{equation}
and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$. However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary  variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.

@@

An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.  
%
\subsection{Datasets and Models}
%
We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different datasets commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with a number of training images in the range $10^4-10^5$.

So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
%
\subsection{Results}
%
Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
%
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-original_samples_ex.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_ex.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
%    \vspace{5mm}
    \begin{subfigure}[t]{0.30\columnwidth}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-glow_samples_ex.pdf}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed c-for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images. Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?

Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom). Despite the fact that the original training images are less diverse as e.g. the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts: 
\begin{itemize}
\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row). Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1. 
\end{itemize}

\begin{figure}
    \centering
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-AB.pdf}\\
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-ATrain.pdf}
	\caption{Evolution of the cosine similarity metric (see text) considering two diffusion-based models trained with two independent datasets of size $N$. On the upper plots are displayed normalized histograms considering $1,000$ generated samples for each of the two models \textit{using as seed the same white noise image}.  On the lower plots, this is the maximum of the "cosine" metric which is considered using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the "cosine" metric of the closest training image from the generated sample). The logarithmic vertical and linear horizontal (cosine value) ranges of each raw are shared among the different histograms. Notice that the horizontal range of the bottom row is restricted to a range closer to the maximum value (i.e. 1) materialized by the red vertical line on each histogram. This figure has been inspired by Figure 2 of \cite{kadkhodaie2024generalization}.}
	\label{fig-cosine-diffusion}
\end{figure}
%
The first fact tell us that clearly the generative model are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset is increasing. These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization}. What is less expected is that we still need $10^5$ samples to be confident that the generation regime applies.

This last remarks invite ourselves to perform the same experiment with the \texttt{Glow} flow-based model with $N=10^5$ trained images. The result is shown on Figure \ref{fig-cosine-glow}. It is quite remarkable that even if the generated samples seems different from any of the trained images\footnote{Also any of horizontal flipped, vertical flipped or horizontal+vertical flipped versions.} as right histogram does not peak at unity cosine similarity, in the same time even seeded by the same white noise image the two models output images are different as the left histogram does not show a peak at unity cosine similarity too. This is clearly a sign that the two models haven't leaned the same forward transport mapping. 
%
\begin{figure}
    \centering
	\includegraphics[width=0.8\textwidth]{fig-glow-cosine-AB_ATrain.pdf}
	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two Glow models using two different datasets of size $N$ but the same as the ones used by the diffusion models. Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image. Right: the cosine similarity of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical).}
	\label{fig-cosine-glow} 
\end{figure}
